---
title: 'STAT 420: Homework 05'
author: "Sahana Epari"
date: 'Due: Tuesday, October 13 by 11:30 PM CT'
output:
  html_document:
    theme: readable
    toc: yes
---

# Directions


# Assignment

## Exercise 1 (Using `lm`)

For this exercise we will use the data stored in [`nutrition.csv`](nutrition.csv). It contains the nutritional values per serving size for a large variety of foods as calculated by the USDA. It is a cleaned version totaling 5,138 observations and is current as of September 2015.

The variables in the dataset are:

- `ID` 
- `Desc` - Short description of food
- `Water` - in grams
- `Calories` 
- `Protein` - in grams
- `Fat` - in grams
- `Carbs` - Carbohydrates, in grams
- `Fiber` - in grams
- `Sugar` - in grams
- `Calcium` - in milligrams
- `Potassium` - in milligrams
- `Sodium` - in milligrams
- `VitaminC` - Vitamin C, in milligrams
- `Chol` - Cholesterol, in milligrams
- `Portion` - Description of standard serving size used in analysis

**(a)** Fit the following multiple linear regression model in `R`. Use `Calories` as the response and `Carbs`, `Fat`, and `Protein` as predictors.

\[
Y_i = \beta_0 + \beta_1 x_{i1} + \beta_2 x_{i2} + \beta_3 x_{i3} + \epsilon_i.
\]

Here,

- $Y_i$ is `Calories`.
- $x_{i1}$ is `Carbs`.
- $x_{i2}$ is `Fat`.
- $x_{i3}$ is `Protein`.

Use an $F$-test to test the significance of the regression. Report the following:
 
- The null and alternative hypotheses
- The value of the test statistic
- The p-value of the test
- A statistical decision at $\alpha = 0.01$
- A conclusion in the context of the problem

When reporting these, you should explicitly state them in your document, not assume that a reader will find and interpret them from a large block of `R` output.

```{r}
nutrition <- read.csv("nutrition.csv")
nutritionmodel <- lm(Calories~Carbs+Fat+Protein,data=nutrition)
nullmodel <- lm(Calories~1,data=nutrition)
anovatest = anova(nullmodel, nutritionmodel)
anovatest
```


$H_O: \beta_1 = \beta_2 = \beta_3 = 0$

$H_1:$ At least one $\beta_1,\beta_2,\beta_3 ≠ 0$

Test Statistic: 152445

P-value: 2.2e-16

Statistical Decision: Since the p-value is less than the alpha value of 0.01, we can reject the null hypothesis.

Conclusion: Using statistical evidence from the ANOVA test, we can say that at least one of the 3 parameters in the model has a linear relationship with Calories.

**(b)** Output only the estimated regression coefficients. Interpret all $\hat{\beta}_j$ coefficients in the context of the problem.

```{r}
summary(nutritionmodel)$coef[,1]
```

$\beta_0=3.768066$: the average number of calories per gram in a type of food with no Carbs, Fat, or Protein.
$\beta_1=3.773605$: the average increase of calories per gram increase in Carbs, keeping Fat and Protein constant.
$\beta_2=8.804109$: the average increase of calories per gram increase in Fat, keeping Carbs and Protein constant.
$\beta_3= 3.967260$: the average increase of calories per gram increase in Protein, keeping Carbs and Fat constant.


**(c)** Use your model to predict the amount of `Calories` in a Big Mac. According to [McDonald's publicized nutrition facts](http://nutrition.mcdonalds.com/getnutrition/nutritionfacts.pdf), the Big Mac contains 47g of carbohydrates, 28g of fat, and 25g of protein.

```{r}
bigmac = data.frame(Carbs=47,Fat=28,Protein=25)
predict(nutritionmodel,newdata=bigmac)
```
According to the model, there should be approximately 526 calories in a Big Mac.

**(d)** Calculate the standard deviation, $s_y$, for the observed values in the Calories variable. Report the value of $s_e$ from your multiple regression model. Interpret both estimates in the context of this problem.

```{r}
sd(nutrition$Calories)
```
```{r}
summary(nutritionmodel)$sigma
```

The $s_y$ is 179.2444, which tells us about the spread of data points around the mean calories value. The $s_e$ is the standard deviation of the residual and tells us how far away the average point is from the model.


**(e)** Report the value of $R^2$ for the model. Interpret its meaning in the context of the problem.

```{r}
summary(nutritionmodel)$r.squared
```

The $R^2$ value is 0.9888987, which tells us that 98.89% of the variance in calories can be attributed to carbs, fats, and proteins. It is indicative of the strength of the model.

**(f)** Calculate a 90% confidence interval for $\beta_2$. Give an interpretation of the interval in the context of the problem.

```{r}
confint(nutritionmodel,level=0.90,"Fat")
```

The 90% CI resides in (8.77893,8.829288), meaning that we are 90% confident that for each gram increase in fat, the average increase in calories is in this interval.

**(g)** Calculate a 95% confidence interval for $\beta_0$. Give an interpretation of the interval in the context of the problem.

```{r}
confint(nutritionmodel,level=0.95,"(Intercept)")
```

The 95% CI resides in (2.802779,4.733353), meaning that we are 95% confident that for a gram of food with no carbs,fats,or protein, the tru mean of calories is in this interval.

**(h)** Use a 99% confidence interval to estimate the mean Calorie content of a small order of McDonald's french fries that has 30g of carbohydrates, 11g of fat, and 2g of protein. Interpret the interval in context.

```{r}
fries = data.frame(Carbs=30,Fat=11,Protein=2)
predict(nutritionmodel,newdata=fries,interval='confidence',level=0.99)
```
 
 The 99% CI resides in (220.8924,222.6195), meaning that we are 99% confident that the true mean of the calories in McDonald's french fries lies within this interval.
 
 
**(i)** Use a 90% prediction interval to predict the Calorie content of new healthy menu item that has 11g of carbohydrates, 1.5g of fat, and 1g of protein. Interpret the interval in context.

```{r}
newfood = data.frame(Carbs=11,Fat=1.5,Protein=1)
predict(nutritionmodel,newdata=newfood,interval='prediction',level=0.90)
```

The 90% CI resides in (31.3649,93.53739), meaning that we are 90% confident that this new healthy food's calories resides within this interval.

## Exercise 2 (More `lm`)

For this exercise we will again use the nutrition data. 

**(a)** Fit a model with Calories as the response and `Carbs`, `Sodium`, `Fat`, and `Protein` as predictors. Use an $F$-test to test the significance of the regression. Report the following:
 
- The null and alternative hypotheses
- The value of the test statistic
- The p-value of the test
- A statistical decision at $\alpha = 0.01$
- A conclusion in the context of the problem

```{r}
nutritionmodel2 <- lm(Calories~Carbs+Sodium+Fat+Protein,data=nutrition)
summary(nutritionmodel2)
```

$H_O: \beta_1 = \beta_2 = \beta_3 = \beta_4 = 0$

$H_1:$ At least one $\beta_1,\beta_2,\beta_3, \beta_4 ≠ 0$

Test Statistic (F): 114343

P-value: 2.2e-16 (approximately 0)

Statistical Decision: Since the p-value is less than the alpha value of 0.01, we can reject the null hypothesis.

Conclusion: Using statistical evidence from the ANOVA test, we can say that at least one of the 4 parameters in the model has a linear relationship with Calories.

**(b)** For each of the predictors in part **(a)**, perform a $t$-test for the significance of its regression coefficient. Report the following for each:
 
- The null and alternative hypotheses
- The value of the test statistic
- The p-value of the test
- A statistical decision at $\alpha = 0.01$

```{r}
summary(nutritionmodel2)$coef
```

**(Carbs:)**

Null hypothesis: $B_1=0$ 

Alternative hypotehsis: $B_1 ≠ 0$

Test Statistic (t): 388.7

P-value: 0

Statistical Decision:  Since the p-value is less than 0.01, we reject the null hypothesis.

**(Sodium:)**

Null hypothesis: $B_2=0$ 

Alternative hypotehsis: $B_2 ≠ 0$

Test Statistic (t): 1.363

P-value: 0.173

Statistical Decision: Since the p-value is greater than 0.01, we fail to reject the null hypothesis.

**(Fat:)**

Null hypothesis: $B_3=0$ 

Alternative hypotehsis: $B_3 ≠ 0$

Test Statistic (t): 575.26

P-value: 0

Statistical Decision: Since the p-value is less than 0.01, we reject the null hypothesis.

**(Protein:)**

Null hypothesis: $B_4=0$ 

Alternative hypotehsis: $B_4 ≠ 0$

Test Statistic (t): 150.53

P-value: 0

Statistical Decision: Since the p-value is less than 0.01, we reject the null hypothesis.

**(c)** Based on your results in part **(b)**, do you still prefer the model in part **(a)**, or is there instead a model with three predictors that you prefer? Briefly explain.

Based on the t-test results, Sodium is the only predictor that concluded there is no relationship between the predictor and number of Calories per gram of food. From this, we can assume that a model that only includes the remaining predictors would be a better option.

## Exercise 3 (Comparing Models)

For this exercise we will use the data stored in [`goalies_cleaned.csv`](goalies_cleaned.csv). It contains career data for 462 players in the National Hockey League who played goaltender at some point up to and including the 2014 - 2015 season. The variables in the dataset are:
 
- `W` - Wins
- `GA` - Goals Against
- `SA` - Shots Against
- `SV` - Saves
- `SV_PCT` - Save Percentage
- `GAA` - Goals Against Average
- `SO` - Shutouts
- `MIN` - Minutes
- `PIM` - Penalties in Minutes
 
**(a)** Fit a multiple linear regression model with Wins as the response and all other variables as the predictors.
 
Use an $F$-test to test the significance of the regression. Report the following:
 
- The null and alternative hypotheses
- The value of the test statistic
- The p-value of the test
- A statistical decision at $\alpha = 0.10$
- A conclusion in the context of the problem
 
When reporting these, you should explicitly state them in your document, not assume that a reader will find and interpret them from a large block of `R` output.

```{r}
goalies <- read.csv("goalies_cleaned.csv")
goaliesmodel = lm(W~.,data=goalies)
summary(goaliesmodel)
```

Null hypothesis: All $B$ values = 0 

Alternative hypotehsis: At least one $B ≠ 0$

Test Statistic (F): 3938

P-value: 2.2e-16 (approx 0)

Statistical Decision:  Since the p-value is less than 0.01, we reject the null hypothesis and conclude that at least one of the predictors has a linear relationship with Wins.

**(b)** Calculate the RMSE of this full model. Report the residual standard error of this full model. What is the relationship of these two values?

Recall, we have defined RMSE as,

\[
RMSE = \sqrt{\frac{1}{n} \sum_{i = 1}^{n}(y_i - \hat{y}_i)^2}.
\]

```{r}
rmse = sqrt(mean(resid(goaliesmodel)^2))
rmse
```

```{r}
summary(goaliesmodel)$sigma
```

The RMSE is 12.40 and the residual standard error is 12.52.

**(c)** Fit a model with Wins as the response and with Goals Against, Goals Against Average, Saves, and Save Percentage as the predictors. Calculate the RMSE of this model.

```{r}
goaliesmodel2 <- lm(W ~ GA + GAA + SV + SV_PCT, data=goalies)
sqrt(mean(resid(goaliesmodel2)^2))
```

**(d)** Fit a model with Wins as the response and with Goals Against Average and Save Percentage as the predictors. Calculate the RMSE of this model.

```{r}
goaliesmodel3 <- lm(W~GAA + SV_PCT, data=goalies)
sqrt(mean(resid(goaliesmodel3)^2))
```

**(e)** Based on the previous three models, which model is most helpful for predicting wins? Briefly explain.

Based on the previously three models, the best one to use is the first model (with all the predictors), because it has the lowest RMSE.

**(f)** Conduct an ANOVA $F$-test comparing the models in parts **(c)** and **(d)**. Report the following:
 
- The null and alternative hypotheses
- The value of the test statistic
- The p-value of the test
- A statistical decision at $\alpha = 0.10$
- A conclusion in the context of the problem
 
When reporting these, you should explicitly state them in your document, not assume that a reader will find and interpret them from a large block of `R` output.

```{r}
anova(goaliesmodel2,goaliesmodel3)
```

Null hypothesis: $B$ values = 0 

Alternative hypotehsis: At least one $B ≠ 0$

Test Statistic (F): 3599.8

P-value: 2.2e-16 (approx 0)

Statistical Decision:  Since the p-value is less than 0.01, we reject the null hypothesis and conclude that one of either goals or saves has a linear relationship with calories and adds to the model.



## Exercise 4 (Regression without `lm`)

For this exercise use the `prostate` dataset from the `faraway` package. Use `?prosate` to learn about the dataset. The goal of this exercise is to fit a model with `lpsa` as the response and the remaining variables as predictors.

**(a)** Obtain the estimated regression coefficients **without** the use of `lm()` or any other built-in functions for regression. That is, you should use only matrix operations. Store the results in a vector `beta_hat_no_lm`. To ensure this is a vector, you may need to use `as.vector()`. Return this vector as well as the results of `sum(beta_hat_no_lm)`.

```{r}
library(faraway)
n = nrow(prostate)
p = 9
x = cbind(rep(1,n), prostate$lcavol, prostate$lweight, prostate$age, prostate$lbph, prostate$svi, prostate$lcp, prostate$gleason, prostate$pgg45)
y = prostate$lpsa

beta_hat_no_lm = as.vector(solve(t(x) %*% x) %*% t(x) %*% y)

beta_hat_no_lm
```

```{r}
sum(beta_hat_no_lm)
```
**(b)** Obtain the estimated regression coefficients **with** the use of `lm()`. Store the results in a vector `beta_hat_lm`. To ensure this is a vector, you may need to use `as.vector()`. Return this vector as well as the results of `sum(beta_hat_lm)`.

```{r}
prostatemodel = lm(lpsa~.,data=prostate)
beta_hat_lm = as.vector(coef(prostatemodel))
beta_hat_lm
```
```{r}
sum(beta_hat_lm)
```
**(c)** Use the `all.equal()` function to verify that the results are the same. You may need to remove the names of one of the vectors. The `as.vector()` function will do this as a side effect, or you can directly use `unname()`.

```{r}
all.equal(beta_hat_lm,beta_hat_no_lm)
```

**(d)** Calculate $s_e$ without the use of `lm()`. That is, continue with your results from **(a)** and perform additional matrix operations to obtain the result. Output this result. Also, verify that this result is the same as the result obtained from `lm()`.

```{r}
y_hat = x %*% solve(t(x) %*% x) %*% t(x) %*% y
e = y - y_hat
se = sqrt((t(e) %*% e) / (n - p))
se
```
```{r}
all.equal(as.vector(se), summary(prostatemodel)$sigma)
```
**(e)** Calculate $R^2$ without the use of `lm()`. That is, continue with your results from **(a)** and **(d)** and perform additional operations to obtain the result. Output this result. Also, verify that this result is the same as the result obtained from `lm()`.

```{r}
SSE = sum(e^2)
SST = sum((y-mean(y))^2)

R_squared = 1 - (SSE/SST)
R_squared
```
```{r}
all.equal(R_squared,summary(prostatemodel)$r.squared)
```
